Cheminformatics: The Computer Science of Chemical Discovery

### 2011-07-12 paper.tex
Final edits

Before the millenium (<year 2000), many chemists were still using paper
lab-notebooks, multiple compound registration, and search tools. Compared to
today there were many pitfalls and drawbacks with this. The overall architecture
was too inflexible to adopt to fast changing data standards, scientists spent
too much time with administrative work, chemical data quality was inconsistent,
an alignment with other working groups was inefficient (e.g. for running
analytical experiments), a legal dispute required to find data in paper
lab-notebooks, data synchronization between systems caused lag times and access
delays, and local silos hampered an efficient global collaboration. Now,
chemists prefer ELNs, because they can run searches in the data of colleagues. This does
not only allow to find experts faster, but allows also access to highly trusted
data with approved quality workflows. Since the data allows also access to
analytical and legal data directly every new ELN release is aiming at increasing
the integration of scientific and business workflows for reducing the risk due
to irrational or bounded-rational decision making and increasing decision making
efficiency.

@@

The future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types, aka compounds. A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures), efficient data synchronization, most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and rely on external services themselves. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity. Efforts like ChEMBL and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring a constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

@@

A number of toolkits are available including the Chemistry Development
Kit (CDK), OpenBabel, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(http://avogadro.openmolecules.net) uses Open Babel, while the
molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
OpenBabel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats.

@@

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and OpenBabel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimization, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developer's mailing list (either cdk-devel@lists.sf.net or
openbabel-devel@lists.sf.net) with their interests and background, and
a mentor will be assigned to help them join the project.

\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of its development from the 1960s until the new millenium
was performed in secret, because of competition in the pharmaceutical
industry.  While chemists were among the first scientists to embrace
web-based tools of knowledge management, research into fundamental
algorithmic questions of structure enumeration and database management
(and a host of other topics) have suffered, because of limited
academic access to either tools or data.  As we enter the second
decade of the 21st century, however, this landscape is rapidly
shifting.  There now exist free-of-charge open-source cheminformatics
toolkits, and open databases with tens of millions of compounds.  We
believe the prospects for academic and industrial collaboration
between chemists and computer scientists are quite bright, and we
would like to encourage more computer scientists to participate in
cheminformatics research.  There is an admittedly steep learning
curve, as the structural issues faced by chemists are often too
detailed and subtle to admit as clean an abstraction as ``algorithms
on a string,'' the way many bioinformatics can be abstracted into a CS
framework.  Nevertheless, because of the changing landscape, there are
many research questions we are only now in position to try to answer:
for a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
open-source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think ``chemically,''
%we may discover ``miracle drugs'' more consistently than ever before.
we are much better positioned taking not only ``simple'' chemicals
into account (only one identifier and isomer possible per molecule),
but also their complex relationships, transformations, and
combinatorial challenges into account. This will allow us to
normalize, search, and use all chemical information, around us and
within us.

>>>>

Before the year 2000, many chemists were still using paper
lab-notebooks, multiple compound registration, and search tools. Compared to
today there were many pitfalls and drawbacks with this. The overall architecture
was too inflexible to adopt to fast changing data standards, scientists spent
too much time with administrative work, chemical data quality was inconsistent,
an alignment with other working groups was inefficient (e.g. for running
analytical experiments), a legal dispute required to find data in paper
lab-notebooks, data synchronization between systems caused lag times and access
delays, and local silos hampered an efficient global collaboration. Now,
chemists prefer ELNs, because they can run searches in the data of colleagues. This does
not only allow to find experts faster, but also allows access to highly trusted
data with approved quality workflows. Since this also gives access to
analytical and legal data directly every new ELN release is aiming at increasing
the integration of scientific and business workflows for reducing the risk due
to irrational or bounded-rational decision making and increasing decision making
efficiency.

@@

The future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types
(compounds). A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures), efficient data synchronization, most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and rely on external services themselves. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity. Efforts like ChEMBL and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring a constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

@@

The extent of open-source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly
due to the existence of a lucrative pharmaceutical market for
proprietary software. However, in recent years there has been a
rise in the amount of open-source cheminformatics software. One
contribution to this has been the Blue Obelisk group
(http://blueobelisk.org) of chemists, who have worked together to
create interoperable open-source chemistry software (among other
activities). Now there are a number of open-source toolkits available
including the Chemistry Development
Kit (CDK), Open Babel, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(http://avogadro.openmolecules.net) uses Open Babel, while the
molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
Open Babel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats.

@@

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimization, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developer's mailing list (either cdk-devel@lists.sf.net or
openbabel-devel@lists.sf.net) with their interests and background, and
a mentor will be assigned to help them join the project.

\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of its development from the 1960s until the new millenium
was performed in secret, because of competition in the pharmaceutical
industry.  While chemists were among the first scientists to embrace
web-based tools of knowledge management, research into fundamental
algorithmic questions of structure enumeration and database management
(and a host of other topics) have suffered, because of limited
academic access to either tools or data.  As we enter the second
decade of the 21st century, however, this landscape is rapidly
shifting.  There now exist free-of-charge open-source cheminformatics
toolkits, and open databases with tens of millions of compounds.  We
believe the prospects for academic and industrial collaboration
between chemists and computer scientists are quite bright, and we
would like to encourage more computer scientists to participate in
cheminformatics research.  There is an admittedly steep learning
curve, as the structural issues faced by chemists are often too
detailed and subtle to admit as clean an abstraction as ``algorithms
on a string,'' the way many bioinformatics algorithms can be abstracted into a CS
framework.  Nevertheless, because of the changing landscape, there are
many research questions we are only now in position to try to answer:
for a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
open-source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think ``chemically,''
%we may discover ``miracle drugs'' more consistently than ever before.
we are much better positioned taking not only ``simple'' chemicals
into account (only one identifier and isomer possible per molecule),
but also their complex relationships, transformations, and
combinatorial challenges into account. This will allow us to
normalize, search, and use all chemical information, around us and
within us.
